Methods that remove batch effects while retaining group differences may lead to exaggerated confidence in downstream analyses

### 2014-11-20 tex/manuscript.tex
rettinger

A simple removal of batch effects can be achieved by subtracting the mean of the measurements in one batch from all measurements in that batch, i.e mean adjustment or one-way ANOVA adjustment as implemented in the method \texttt{pamr.batchadjust} from the \texttt{pamr} package in R. When the batch--group design is balanced, mean-adjustment will remove most, but not necessarily all, variance attributed to batch and leave the between group variance, thus increasing the statistical power. However, when the batch--group design is unbalanced, batch differences will in part be influenced by group differences, and thus batch correction will reduce group differences and thereby reduce the statistical power. In very uneven group--batch designs with multiple groups, spurious group differences may even be induced in this way. Figure~\ref{fig:boxplots}c illustrates both of these effects.

@@

A popular tool for batch adjustment of gene expression data is ComBat (\citealp{Johnson2007}) which is now included in the R bioconductor package \texttt{sva} (\citealp{Leek2012}). This allows the inclusion of covariates, e.g. group difference, the effects of which should not be removed in the batch adjustment. In addition, it uses an empirical Bayes approach to avoid over-correction for batch effects, which is critical for small batches. Increasingly, the inclusion of group difference as a covariate when removing batch effects has been recommended, both in the \texttt{sva} tutorial and user fora. ComBat has thus helped popularise this approach. Based on the actual use of batch adjustment methods like ComBat by the authors and others, ourselves included, we suspect that adjusted data are commonly treated as ``batch effect free'' in subsequent analyses, and as a consequence, confidence in group effects has been overestimated and false results reported. The commercial software Partek \CITE{citation or reference needed?}, and the R packages \texttt{limma} (\citealp{Smyth2003}) and \texttt{ber} (\citealp{Giordan2013}) also offer batch adjustment with covariates. Seeing that \texttt{ber} is sparsly used, and \texttt{Partek} and the relevant method ``removeBatchEffect'' in \texttt{limma} note that the this is not intended used prior to linear modeling, we choose to use the more popular and well documented ComBat to demonstrate the problem. We note that the \texttt{sva} are now being updated to address these concerns.

@@

The most common method for removing batch effects is to zero-centre each batch:
\begin{equation}
\tilde{Y}^0_{ijr}=Y_{ijr}-\bar{Y}_{i}
\quad\text{where}\quad
\bar{Y}_i=\frac{1}{n_{i\textrm{-}}} \sum_{j=1}^M\sum_{r=1}^{n_{ij}} Y_{ijr},\;
n_{i\textrm{-}}=\sum_{j=1}^M n_{ij}.
\end{equation}
An alternative is to centre each batch to the common average by adding the average value $\bar{Y}$ across the entire data set: i.e. $\tilde{Y}^{\textrm{avg}}_{ijr}=\tilde{Y}^0_{ijr}+\bar{Y}$. When comparing groups, the common value $\bar{Y}$ has no effect, and so this is equivalent to zero-centring each batch. If the groups are unevenly represented in the different batches, the batch average $\bar{Y}_i$ will tend to capture, through $\bar\beta_i$, group differences, as well as batch effects:
\begin{equation}
\tilde{Y}^0_{ijr}=\beta_j-\bar{\beta}_i+\epsilon_{ijr}-\bar{\epsilon}_i
\quad\text{where}\quad
\bar{\beta}_i={\sum_{j=1}^M \frac{n_{ij}}{n_{i\textrm{-}}}\beta_j},\;
\bar{\epsilon}_i=\frac{1}{n_{i\textrm{-}}} \sum_{j=1}^M\sum_{r=1}^{n_{ij}} \epsilon_{ijr}.
\end{equation}
Thus, batch centering will tend to reduce group differences in an unbalanced design, and reduce the power of downstream analyses. By reducing the differences between some groups, i.e. those found together in the same batches, one may also induce false differences between other groups, as is demonstrated in Figure~\ref{fig:boxplots}c.

@@

The undesired consequences of preserving group effects when correcting for batch effect is readily illustrated with a sanity check using random numbers. The documentation accompanying the \texttt{sva} library has an executable example demonstrating how to adjust a data set with ComBat, followed by an F-test. Replacing the real data with random numbers from a standard normal distribution, $\text{N}(0,1)$, but otherwise following the instructions, will generate the p-value distribution shown in Figure~\ref{fig:sanity}a.

As can be seen from the QQ plot in Figure~\ref{fig:sanity}b, the main effect of the procedure is to inflate the F-statistic by a factor. The size of this factor depends on how unbalanced the group--batch design is, rather than on the sample size in itself. Thus, increasing the sample size will not reduce the problem.

@@

In an experiment described in \citet{Towfic2014}, the effect of glatiramer acetate (a medicine for multiple sclerosis) was compared to the effect of a generic drug. Cells were treated with glatiramer acetate (34 samples), the generic drug (11 samples), or one of 14 other treatments (64 samples), and mRNA was measured using microarrays. A batch effect correlating to the chip (Illumina WG-6\_V2, six samples per chip, 17 chips in total) was observed and adjusted for with ComBat using treatment as a covariate. Batch adjusted data were then tested for differentially expressed genes, yielding hundreds of differentially expressed probes (Table S5, \citealp{Towfic2014}) using a $\text{FDR}<0.05$ threshold. Unfortunately, the batch--treatment design was highly unbalanced, with several batches having only one of the main treatments of interest. We re-analyzed the cited data (GEO: GSE40566) without using ComBat, but instead blocked for batch effect in \texttt{limma}, detecting only 9 differentially expressed genes at $\text{FDR}<0.05$. The sanity check  outlined above with random numbers was also carried out. The distribution of p-values for different settings are shown in Figure~\ref{fig:comparison}a.

@@

The supporting information of the original ComBat article (\citealp{Johnson2007}) demonstrates the method on cells inhibited for the expression of the TAL1 gene compared to controls on a microarray platform (denoted ``Data set 2''). The experiment consist of 30 samples in 3 batches (batch1:6/2, batch2:3/4 and batch3:9/6 treatment/control samples). ComBat was applied followed by a T-test, in order to identify differentially expressed genes. First, we reproduced their analysis, including the adjustment by ComBat, but using \texttt{limma} instead of the T-test, resulting in 1003 probes ($q<0.05$).  Then, we analysed their data without batch adjustment in ComBat, but blocking for batch in \texttt{limma}, resulting in 377 probes ($q<0.05$). In addition, the sanity checks outlined above were performed. The distribution of P-values for different settings are shown in Figure~\ref{fig:comparison}b. In contrast to the results obtained for \citet{Towfic2014} (Figure~\ref{fig:comparison}a), the P-value distributions for the alternative analysis do not indicate a huge difference. Nevertheless, the P-values may still be somewhat deflated in the ComBat adjusted analysis.

@@

In the original ComBat article (\citealp{Johnson2007}), the main objective was to employ an empirical Bayes method to allow better handling of small batches. The inclusion of covariates, e.g. for retaining group differences for unbalanced designs, was optional and appeared subordinate, only exemplified in the supplementary information. However, over the years, this feature became more important, judging from advice given on user fora.
Later, when ComBat was incorporated into the \texttt{sva} package (\citealp{Leek2012}), specification of a covariate model became required: batch adjustment without covariates was no longer a default behaviour. Although specification of a null model was still possible, the help file specifies that covariates should contain ``outcome of interest and other covariates besides batch''. Thus, we expect this usage to have grown more common over time.

@@

These problems are not specific to ComBat, but are shared by other batch adjustment methods: indeed, the empirical Bayes approach used by ComBat  will dampen the batch adjustments and may thus reduce the problem slightly relative to a more direct two-way ANOVA approach. Although some of the tools warn against performing subsequent analyses on batch adjusted data, these warnings are often somewhat vague and seem to be primarily concerned about the degrees of freedom in the data set. Batch adjustment will reduce the degrees of freedom in the data which may influence the distribution of the test statistics in subsequent analyses, e.g. $F$-statistics from ANOVA. For small data sets or where there are many small batches this may be a problem, while it would be less of a concern when there are a fair number of samples in each batch. However, we find that the main problem is not related to the degrees of freedom, but estimation errors during the batch adjustments which get applied across the data, which can induce group differences in the adjusted data set even in the absence of batch effects.

@@

We are grateful to Geir Kjetil Sandve for useful suggestions and discussions.

This work was supported by the EUROCAN platform (VN) and the MetAction project (EAR).

@@

\begin{figure}[!p]
\begin{center}$
\begin{array}{cc}
\includegraphics[width=6.5cm]{Fig/boxplots_v3_meanadjusted.pdf} &
\includegraphics[width=6.5cm]{Fig/boxplots_v3_anovaadjusted.pdf}
\end{array}$
\end{center}
\caption
{
We simulated expression of one gene from two and three study groups. These two(a-d and e-h) cases, designs and effects, were selected to illustrate the spurious effects that may arise from different batch adjustments. The Y-axis represents the log expression values. while the X-axis is used to separate the samples. Circles, triangles, and crosses indicate sample measurements from separate groups. Horizontal dashed lines indicates the mean of the expression values for that groups samples. Vertical dotted lines separates batches with the batch numbers on the lower half of the plot. To the far right of the plots, a typical differential expression test is performed with estimated fold change and a p-value. The mean expression for each group and 95\% confidence interval is estimated and depicted as boxes.
a) Simulated expression of one gene in 26 samples from 3 groups (circles, triangles and crosses). The fold change and p-value presented are from the circles vs. triangles comparison and calculated with a one-way ANOVA. The confidence intervals indicate that the crosses are different from the other two groups.
b) The values from (a) is divided into 2 batches with batch 1 containing circles and triangles, and batch 2 containing triangles and crosses. Batch effects were added, -1 for batch 1 and +1 for batch 2. The ANOVA test now results in a p<.01 between the circles and triangles. This plot illustrates the problem of disregarding batch effects.
c) The values in (b) were batch adjusted using the mean adjustment method. In order to make the means of the two batches equal, some of the group-effect for the crosses were transferred over to the triangels within the same batch. As a result the ANOVA test now indicate a difference between the circle-group and the triangle-group (p<.01).
d) The same values as in (b) are shown and used in a two-way ANOVA with group and batch included in the model. The estimated fold change and corresponding p-value is in effect calculated from the samples in batch 1.
e) Simulated expression of one gene in 20 samples from two groups (circles and triangles). The t-test indicates no significant group differences(p=.82).
f) The values from (e) are divided into 3 batches in an unbalanced way, with only batch 2 containing samples from both groups. Batch effects were added, +1 for batch 1 and -1 for batch 3. The t-test now results in p<.01. This plot illustrates the problem with disregarding batch effects.
g) The values in (f) were batch effect adjusted while retaining the group difference (using removeBatchEffects), thus creating a ``batch effect free'' data set. Notice that all the values in the "circle only" batch 1 are adjusted to make the group mean coincide with the mean of the two circle-samples in batch 2. The triangle-samples in batch 3 are adjusted similarly to match the mean of the two triangle-sampels in batch 2.The t-test now results in a significant group difference (p=.01) and a fold change equal to a fold change calculated from the batch 2 samples only. Thus, the fold change is in effect estimated from only 2+2 samples, but the t-test assumes 10+10 samples with the low p-value as a result.
h) The same values as in (f) are shown and used in a two-way ANOVA with group and batch included in the model. The estimated fold change is again in effect only based on samples from batch 2 and is exactly the same as for (g), but the p-value is much higher and the confidence intervals are more appropriate.
}
\label{fig:boxplots_v3}
\end{figure}

>>>>

A simple removal of batch effects can be achieved by subtracting the mean of the measurements in one batch from all measurements in that batch, i.e zero-centering or one-way ANOVA adjustment as implemented in the method \texttt{pamr.batchadjust} from the \texttt{pamr} package in R. When the batch--group design is balanced, zero-centering will remove most, but not necessarily all, variance attributed to batch and leave the between group variance, thus increasing the statistical power. However, when the batch--group design is unbalanced, batch differences will in part be influenced by group differences, and thus batch correction will reduce group differences and thereby reduce the statistical power. In very uneven group--batch designs with multiple groups, spurious group differences may even be induced in this way. Figure~\ref{fig:boxplots}c illustrates both of these effects.

@@

A popular tool for batch adjustment of gene expression data is ComBat (\citealp{Johnson2007}) which is now included in the R bioconductor package \texttt{sva} (\citealp{Leek2012}). This allows the inclusion of covariates, e.g. group difference, the effects of which should not be removed in the batch adjustment. In addition, it uses an empirical Bayes approach to avoid over-correction for batch effects, which is critical for small batches. Increasingly, the inclusion of group difference as a covariate when removing batch effects has been recommended, both in the \texttt{sva} tutorial and user fora. ComBat has thus helped popularise this approach. Based on the actual use of batch adjustment methods like ComBat by the authors and others, ourselves included, we suspect that adjusted data are commonly treated as ``batch effect free'' in subsequent analyses, and as a consequence, confidence in group effects has been overestimated and false results reported. The commercial software Partek Genomics Suite, and the R packages \texttt{limma} (\citealp{Smyth2003}) and \texttt{ber} (\citealp{Giordan2013}) also offer batch adjustment with covariates. Seeing that \texttt{ber} is sparsly used, and \texttt{Partek} and the relevant method ``removeBatchEffect'' in \texttt{limma} note that this is not intended used prior to linear modeling, we choose to use the more popular and well documented ComBat to demonstrate the problem. We note that the \texttt{sva} are now being updated to address these concerns.

@@

The most common method for removing batch effects is to zero-centre each batch:
\begin{equation}
\tilde{Y}^0_{ijr}=Y_{ijr}-\bar{Y}_{i}
\quad\text{where}\quad
\bar{Y}_i=\frac{1}{n_{i\textrm{-}}} \sum_{j=1}^M\sum_{r=1}^{n_{ij}} Y_{ijr},\;
n_{i\textrm{-}}=\sum_{j=1}^M n_{ij}.
\end{equation}
An alternative is to centre each batch to the common average by adding the average value $\bar{Y}$ across the entire data set: i.e. $\tilde{Y}^{\textrm{avg}}_{ijr}=\tilde{Y}^0_{ijr}+\bar{Y}$. When comparing groups, the common value $\bar{Y}$ has no effect, and so this mean-centring is equivalent to zero-centring each batch. If the groups are unevenly represented in the different batches, the batch average $\bar{Y}_i$ will tend to capture, through $\bar\beta_i$, group differences, as well as batch effects:
\begin{equation}
\tilde{Y}^0_{ijr}=\beta_j-\bar{\beta}_i+\epsilon_{ijr}-\bar{\epsilon}_i
\quad\text{where}\quad
\bar{\beta}_i={\sum_{j=1}^M \frac{n_{ij}}{n_{i\textrm{-}}}\beta_j},\;
\bar{\epsilon}_i=\frac{1}{n_{i\textrm{-}}} \sum_{j=1}^M\sum_{r=1}^{n_{ij}} \epsilon_{ijr}.
\end{equation}
Thus, batch centering will tend to reduce group differences in an unbalanced design, and reduce the power of downstream analyses. By reducing the differences between some groups, i.e. those found together in the same batches, one may also induce false differences between other groups, as is demonstrated in Figure~\ref{fig:boxplots}c.

@@

The undesired consequences of preserving group effects when correcting for batch effect is readily illustrated with a sanity check using random numbers. The documentation accompanying the \texttt{sva} library (v3.8.0) has an executable example demonstrating how to adjust a data set with ComBat, followed by an F-test. Replacing the real data with random numbers from a standard normal distribution, $\text{N}(0,1)$, but otherwise following the instructions, will generate the p-value distribution shown in Figure~\ref{fig:sanity}a.

As can be seen from the QQ plot in Figure~\ref{fig:sanity}b, the main effect of the procedure is to inflate the F-statistic by a factor. The size of this factor depends on how unbalanced the group--batch design is, rather than on the sample size in itself. Thus, increasing the sample size will not reduce the problem as shown in Figure~\ref{fig:sanity}c.

@@

In an experiment described in \citet{Towfic2014}, the effect of glatiramer acetate (a medicine for multiple sclerosis) was compared to the effect of a generic drug. Cells were treated with glatiramer acetate (34 samples), the generic drug (11 samples), or one of 14 other treatments (64 samples), and mRNA expression was measured using microarrays. A batch effect correlating to the chip (Illumina WG-6\_V2, six samples per chip, 17 chips in total) was observed and adjusted for with ComBat using treatment as a covariate. Batch adjusted data were then tested for differentially expressed genes, yielding hundreds of differentially expressed probes (Table S5, \citealp{Towfic2014}) using a $\text{FDR}<0.05$ threshold. Unfortunately, the batch--treatment design was highly unbalanced, with several batches having only one of the main treatments of interest. We re-analyzed the cited data (GEO: GSE40566) without using ComBat, but instead blocked for batch effect in \texttt{limma}, detecting only 11 differentially expressed genes at $\text{FDR}<0.05$. The sanity check using random numbers, as described in section 3.1, was also carried out. The distribution of p-values for different settings are shown in Figure~\ref{fig:comparison}a.

@@

The supporting information of the original ComBat article (\citealp{Johnson2007}) demonstrates the method on cells inhibited for the expression of the TAL1 gene compared to controls on a microarray platform (denoted ``Data set 2''). The experiment consist of 30 samples in 3 batches (batch1:6/2, batch2:3/4 and batch3:9/6 treatment/control samples). ComBat was applied followed by a T-test, in order to identify differentially expressed genes. First, we reproduced their analysis, including the adjustment by ComBat, but using \texttt{limma} instead of the T-test, resulting in 1003 probes ($q<0.05$).  Then, we analysed their data without batch adjustment in ComBat, but blocking for batch in \texttt{limma}, resulting in 377 probes ($q<0.05$). In addition, the sanity check using random numbers was performed. The distribution of P-values for different settings are shown in Figure~\ref{fig:comparison}b. In contrast to the results obtained for \citet{Towfic2014} (Figure~\ref{fig:comparison}a), the P-value distributions for the alternative analysis do not indicate a huge difference. Nevertheless, the P-values may still be somewhat deflated in the ComBat adjusted analysis.

@@

In the original ComBat article (\citealp{Johnson2007}), the main objective was to employ an empirical Bayes method to allow better handling of small batches. The inclusion of covariates, e.g. for retaining group differences for unbalanced designs, was optional and appeared subordinate, only exemplified in the supplementary information. However, over the years, this feature became more important, judging from advice given on user fora.
Later, when ComBat was incorporated into the \texttt{sva} package (\citealp{Leek2012}), specification of a covariate model became required: batch adjustment without covariates was no longer a default behaviour. Although specification of an empty model was still possible, the help file specifies that covariates should contain ``outcome of interest and other covariates besides batch''. Thus, we expect this usage to have grown more common over time.

@@

Problems with batch adjusting unbalanced data sets are not specific to ComBat, but are shared by other batch adjustment methods: indeed, the empirical Bayes approach used by ComBat  will dampen the batch adjustments and may thus reduce the problem slightly relative to a more direct two-way ANOVA approach. Although some of the tools warn against performing subsequent analyses on batch adjusted data, these warnings are often somewhat vague and seem to be primarily concerned about the degrees of freedom in the data set. Batch adjustment will reduce the degrees of freedom in the data which may influence the distribution of the test statistics in subsequent analyses, e.g. $F$-statistics from ANOVA. For small data sets or where there are many small batches this may be a problem, while it would be less of a concern when there are a fair number of samples in each batch. However, we find that the main problem is not related to the degrees of freedom, but estimation errors during the batch adjustments which get applied across the data, which can induce group differences in the adjusted data set even in the absence of batch effects.

@@

We would like to thank Ben Zeskind and coauthors of \citet{Towfic2014} for their responsiveness and sharing scripts and data.
We are also grateful to Geir Kjetil Sandve for useful suggestions and discussions.

This work was supported by the EUROCAN platform (VN, FP7/2007-2013, grant 260791) and the MetAction project (EAR).

@@
